Ensuring Fairness with Transparent Auditing of Quantitative Bias in AI Systems

### 2024-12-12 slides/2024_12_13.tex
Added Dec 13rd slides

\newdate{date}{30}{08}{2024}
\date{\displaydate{date}}
\title[Fairness]{Ensuring Fairness with Transparent Auditing of
Quantitative Bias in AI Systems}
\author[Rex]{
    \begin{CJK}{UTF8}{bsmi}袁至誠\end{CJK}\newline
    Chih-cheng Rex Yuan\newline
    \href{https://rexyuan.com/}{rexyuan.com}
    }
\institute[IIS,AS]{Institute of Information Science, Academia Sinica}

@@

\begin{frame}
    \frametitle{Abstraction}
    Our framework tool abstracts the fairness measures:
    \begin{itemize}
        \item A \emph{row} is a lookup table or dictionary.
        $r_n(\text{``sex''}) = \text{``Female''}$ means $r_n$'s sex is female.
        \item A \emph{privileged predicate} $\priv$ takes a row and determines if
        it belongs to the privileged group.
        For example, $\priv(r_i) := r_i(\text{``race''}) == \text{``Caucasian''}$
        means the privileged group is those with race being Caucasian.
        \item A \emph{positive predicate} $\pos$ takes a row and determines if its
        prediction is positive. For example,
        $\pos(r_i) := r_i(\text{``score''}) > 7$ means a
        row's prediction is positive if its score is greater than 7.
        \item A \emph{ground truth predicate} $\tru$ takes a row and gives the ground
        truth of the result. For example,
        $\tru(r_i) := r_i(\text{``recid''}) == \text{True}$ indicates a
        row's actual recidivism.
    \end{itemize}
\end{frame}

@@

\begin{frame}[fragile]
    \frametitle{Definition}
    \begin{align*}
        \text{equal\_opportunity} (\epsilon, \priv, \pos, \tru)
    \end{align*}

    This is implemented in our tool as a Python function:

    \begin{center}
    \begin{minipage}{0.8\textwidth}
    \begin{minted}[fontsize=\small, frame=single, linenos]{python}
def equal_opportunity(
    ratio,
    privileged_predicate,
    positive_predicate,
    truth_predicate):
    ...
    \end{minted}
    \end{minipage}
    \end{center}
\end{frame}

\begin{frame}
    \frametitle{Application}
    For demonstration, we analyzed the COMPAS dataset.
    Let's set the predicates:
    \begin{align*}
        \priv(r_i) & := r_i(\text{``race''}) \neq \text{``African-American''}\\
        \pos(r_i)  & := r_i(\text{``score\_text''}) \in \set{\text{``Medium''},\text{``High''}}\\
        \tru(r_i)  & := r_i(\text{``two\_year\_recid''}) == \text{True}
    \end{align*}
    These predicates can be published for transparency.
\end{frame}

\begin{frame}[fragile]
    \frametitle{Application}
    \begin{align*}
        \priv(r_i) & := r_i(\text{``race''}) \neq \text{``African-American''}\\
        \pos(r_i)  & := r_i(\text{``score\_text''}) \in \set{\text{``Medium''},\text{``High''}}\\
        \tru(r_i)  & := r_i(\text{``two\_year\_recid''}) == \text{True}
    \end{align*}
    These predicates can then be encoded with Python functions:
    \begin{center}
    \begin{minipage}{0.8\textwidth}
    \begin{minted}[fontsize=\small, frame=single, linenos]{python}
equal_opportunity(
    0.2,
    lambda row: row['race'] != 'African-American',
    lambda row: row['score_text']
      in {'Medium', 'High'},
    lambda row: row['is_recid'] == '1')
    \end{minted}
    \end{minipage}
    \end{center}
\end{frame}

@@

\begin{frame}
    \frametitle{Application}
    \centering
    \includegraphics[width=0.73\textwidth]{Asian}
    \captionof{figure}{Unprivileged Group: Asian (32 rows)}
\end{frame}

\begin{frame}
    \frametitle{Conclusion}
    \begin{itemize}
        \item Decision making by AI may be biased.
        \item With our framework tool, auditors can comprehensively
        review the fairness of an AI system.
    \end{itemize}

>>>>

\newdate{date}{13}{12}{2024}
\date{\displaydate{date}}
\title[AI Fairness]{Ensuring Fairness with Transparent Auditing of
Quantitative Bias in AI Systems}
\author[Rex]{
    \begin{CJK}{UTF8}{bsmi}袁至誠\end{CJK}\newline
    Chih-cheng Rex Yuan\newline
    \href{https://rexyuan.com/}{rexyuan.com}
    }
\institute[IIS,AS]{Institute of Information Science, Academia Sinica}

@@

\begin{frame}
    \frametitle{Abstraction}
    Our framework tool abstracts the fairness measures:
    \begin{itemize}
        \item A \emph{row} is a lookup table or dictionary.
        $r_n(\text{``sex''}) = \text{``Female''}$ means $r_n$'s sex is female.
        \item A \emph{privileged predicate} $\priv$ takes a row and determines if
        it belongs to the privileged group.
        For example, $\priv(r_i) := r_i(\text{``race''}) = \text{``Caucasian''}$
        means the privileged group is those with race being Caucasian.
        \item A \emph{positive predicate} $\pos$ takes a row and determines if its
        prediction is positive. For example,
        $\pos(r_i) := r_i(\text{``score''}) > 7$ means a
        row's prediction is positive if its score is greater than 7.
        \item A \emph{ground truth predicate} $\tru$ takes a row and gives the ground
        truth of the result. For example,
        $\tru(r_i) := r_i(\text{``recid''}) = \text{True}$ indicates a
        row's actual recidivism.
    \end{itemize}
\end{frame}

@@

\begin{frame}
    \frametitle{Application}
    For demonstration, we analyzed the COMPAS dataset.
    Let's set the predicates:
    \begin{align*}
        \priv(r_i) & := r_i(\text{``race''}) \neq \text{``African-American''}\\
        \pos(r_i)  & := r_i(\text{``score\_text''}) \in \set{\text{``Medium''},\text{``High''}}\\
        \tru(r_i)  & := r_i(\text{``two\_year\_recid''}) = \text{True}
    \end{align*}
    These predicates can be published for transparency.
\end{frame}

@@

\begin{frame}
    \frametitle{Conclusion}
    \begin{itemize}
        \item Decision making by AI may be biased.
        \item With our framework and tool, auditors can comprehensively
        review the fairness of an AI system.
    \end{itemize}
